---
title: "A personalized and non-parametric framework for detecting changes in gait cycles"
author:
  - name: "Saeb Ragani Lamooki ^[Email: saebraga@buffalo.edu | Website: <a href=\"https://www.linkedin.com/in/saeb-ragani-lamooki-123a9658/\">LinkedIn</a>]"
    affiliation: Department of Mechanical & Aerospace Engineering, University at Buffalo
  - name: "Jiyeon Kang ^[Email: jiyeonk@buffalo.edu | Phone: +1-716-645-6063 | Website: <a href=\"http://engineering.buffalo.edu/industrial-systems/people/faculty-directory/j-jang.html\">University at Buffalo Official</a>]"
    affiliation: Department of Mechanical & Aerospace Engineering, University at Buffalo
  - name: "Lora A. Cavuoto ^[Email: loracavu@buffalo.edu | Phone: +1-716-645-4696 | Website: <a href=\"http://engineering.buffalo.edu/industrial-systems/people/faculty-directory/cavuoto-lora.html\">University at Buffalo Official</a>]"
    affiliation: Department of Industrial and Systems Engineering, University at Buffalo
  - name: "Fadel M. Megahed ^[Email: fmegahed@miamioh.edu | Phone: +1-513-529-4185 | Website: <a href=\"https://miamioh.edu/fsb/directory/?up=/directory/megahefm\">Miami University Official</a>]"
    affiliation: Farmer School of Business, Miami University
  - name: "Allison Jones Farmer ^[Email: farmerl2@miamioh.edu | Phone: +1-513-529-4823 | Website: <a href=\"https://miamioh.edu/fsb/directory/?up=/directory/farmerl2\">Miami University Official</a>]"
    affiliation: Farmer School of Business, Miami University
bibliography: gaitRefs.bib
csl: apa.csl
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
    number_sections: TRUE
    theme: simplex
    paged_df: TRUE
    code_folding: show
  includes:
    in_header: structure.tex
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      cache = TRUE,
                      progress = FALSE, 
                      verbose = FALSE,
                      dpi = 600)
options(qwraps2_markup = "markdown")


```




---

# R Setup and Required Packages
In this project, the open-source R programming language is used to model/monitor changes in the gait cycles over time. R is maintained by an international team of developers who make the language available at [The Comprehensive R Archive Network](https://cran.r-project.org/). Readers interested in reusing our code and reproducing our results should have R installed locally on their machines. R can be installed on a number of different operating systems (see [Windows](https://cran.r-project.org/bin/windows/), [Mac](https://cran.r-project.org/bin/macosx/), and [Linux](https://cran.r-project.org/bin/linux/) for the installation instructions for these systems). We also recommend using the RStudio interface for R. The reader can [download RStudio](http://www.rstudio.com/ide) for free by following the instructions at the link. For non-R users, we recommend the [Hands-on Programming with R](https://rstudio-education.github.io/hopr/packages.html) for a brief overview of the software's functionality. Hereafter, we assume that the reader has an introductory understanding of the R programming language.

In the code chunk below, we load the packages used to support our analysis.  

```{r packages, cache=FALSE}
if(require(checkpoint)==FALSE) install.packages("checkpoint") # check to see if checkpoint is installed; if not, install it
library(checkpoint) # package used to facilitate the reproducibility of our work

# a checkpoint of R packages on CRAN on June 08, 2020 to enable the reproduction of our work in the future
checkpoint("2020-06-08")

# check if packages are not installed; if yes, install missing packages
packages = c("tidyverse", "magrittr", # typical data analysis packages
             "foreach", "parallel", # packages used for parallelizing some code chunks
             "MALDIquant", # match closest points between two vectors
             "MASS", # calculates the generalized inverse matrix
             "fda.usc", # Used for visual diagnostic plotting
             "R.matlab")
newPackages = packages[!(packages %in% installed.packages()[,"Package"])]
if(length(newPackages) > 0) install.packages(newPackages)

# using the library command to load all packages; invisible used to avoid printing all packages and dependencies used
invisible(lapply(packages, library, character.only = TRUE))

source("./Functions.R") # our custom built functions

set.seed(2020)
startTime <- Sys.time()
```


---

# Extracting, Transforming and Loading (ETL) the Gait Data into R
In this paper, we perform a secondary data analysis of the experimental data reported in @baghdadi2019monitoring, where fifteen participants were equipped with an IMU sensor wrapped on their right ankle. We used the IMU data to transform the linear acceleration signals from the local frame of reference to the global frame of reference. Next, we calculated the acceleration magnitude signal and smoothed it by applying a low pass butterworth filter with a cut-off frequency of 10 Hz. Further, we segmented the idividual gait cycles from foot-flat to foot-flat for each participant by tuning the parameters of the segmentation algorithm for each participant.  

We load **four different sets of data** into R:

  -  *Subject heights*, which is read directly from the [GitHub Repository](https://github.com/fmegahed/fatigue-changepoint) of the @baghdadi2019monitoring paper. Note that we have excluded Subject 13's data as the segmentation algorithm falls short of detecting the gaits in this subject. Shape of acceleration magnitude signal did not contain the distinguishable larger peak, commonly reflected in gait cycles. Moreover, we did not include the results for subjects 5 & 7 in our analysis as they did not complete the experiment. This inclusion criteria was considered to allow for enough data to construct the 3 phases effectively.
  - *Features*, where we extract the positions in Frennet Serret coordinates as well as the stride duration. *Stride length*, *stride height*, and *stride duration* along with *experimental time* for all subjects are subsequently calculated. The $1^{st}$ three are important kinematic features that are used to model/monitor changes in gait in one of the two schemes, implemented in this work. Note that we did not use this data directly from the @baghdadi2019monitoring paper since our segmentation here was from foot-flat to foot-flat and a different cut-off frequency for the Butterworth filter was used (given that the goals of our paper and the original paper are different).    
  - *Profiles*, where the points forming each gait segment are mantained. The profiles present an alternative approach to monitoring and modeling changes in the gait cycles.   
  - *Video tags*, contains the time stamps when subjects enter and leave the main laboratory space where the non-walking task (loading/unloading) takes place. This information is used to ensure the consistency of the task being performed by the participant, i.e., we do not want to detect the changes due to a change in the undertaken task, but rather due to fatigue (see the discussion in our manuscript for more details).

The code chunk below provides our code for reading the three different datasets. The *subject heights* dataset is saved into a vector, which we use to scale the computed gait/stride length and height. The main outputs from this chunk are two lists (a) **subjectFeaturesList** which contains the *scaled stride length*, *scaled stride height*, *stride duration* and *experimental time*, and (b) **subjectProfilesList** which contains the sample IMU acceleration magnitude points making each acceleration profile.
```{r read}
# Reading the subject Heights from the Baghdadi et al. (2019) GitHub Repo:
# ------------------------------------------------------------------------
subject_heights <- read.csv("https://raw.githubusercontent.com/fmegahed/fatigue-changepoint/master/Data/Raw/DemographicData.csv") %>%   pull(Height) %>% `*` (0.01) %>%  .[-13] # pull height and convert into meters

# Computing gait features based on experimental data from Baghdadi et al. (2019):
# -------------------------------------------------------------------------------
featureFiles <- list.files(path = "../Data/matFiles/Features/", pattern = "*.mat", full.names = TRUE)
rawSubjectsData <- lapply(featureFiles, readMat) %>% # read all mat files in Feature List
  lapply("[[", "M.i.k") # extract sublist titled "M.i.k" from all the lists
names(rawSubjectsData) <- str_extract_all(featureFiles, "Subject[:digit:]*", simplify = TRUE)

numRows <- lengths(rawSubjectsData)/17 # number of observations per Subject

subjectFeaturesList <- vector(mode = "list", length = length(rawSubjectsData)) # initialize list
names(subjectFeaturesList) <- names(rawSubjectsData)

for (i in seq(length(rawSubjectsData))) {
  StrideLength <-  rawSubjectsData[[i]][ 1:numRows[i] ] %>% 
    lapply(range) %>% lapply(diff) %>% unlist() %>% 
    `/` (subject_heights[i])
  
  StrideHeight <- rawSubjectsData[[i]][ (1+numRows[i]):(2*numRows[i]) ] %>% 
    lapply(range) %>% lapply(diff) %>% unlist() %>% 
    `/` (subject_heights[i])
  
  StrideDuration <- rawSubjectsData[[i]][ (1+ 2*numRows[i]):(3*numRows[i]) ] %>% 
    lapply(range) %>% lapply(diff) %>% unlist(recursive = TRUE)
  
  expTime <- rawSubjectsData[[i]][ (1+ 16*numRows[i]):(17*numRows[i]) ] %>% 
    lapply(unlist) %>% lapply(min) %>% unlist()
  
  subjectFeaturesList[[i]] <- cbind(StrideLength, StrideHeight, StrideDuration, expTime)
}


# Reading the segmented gait cycles for each participant:
# -------------------------------------------------------
profileFiles <- list.files(path = "../Data/matFiles/Profiles/", pattern = "*.mat", full.names = TRUE) 
subjectProfilesList <- lapply(profileFiles, readMat) %>% # read all mat files from the path above
  lapply("[[", "gait") # extract sublist titled "gait" from all the lists
names(subjectProfilesList) <- sapply(profileFiles, str_extract_all, "Subject[:digit:]*", simplify = TRUE)

# Reading task-related based on the tagging of the experimental videos:
# ---------------------------------------------------------------------
videoFiles <- list.files(path = "../Data/csvFiles/", pattern = "*.csv", full.names = TRUE) 
videoDataList <- lapply(videoFiles, read.csv)
names(videoDataList) <- sapply(videoFiles, str_extract_all, "Subject[:digit:]*", simplify = TRUE)

```

---

# Data Preprocessing and Rational Subgroups

## Data Preprocessing
In Section 2, we have read four different data sets and extracted the three gait features. In this subsection, we perform further preprocessing of the data. Specifically, we remove the first 10 minutes of experimental data from the `videoDataList` since these corresponded to the warm-up period of the experiment as detailed in @maman2017data, @baghdadi2019monitoring, and @maman2020data. Next, we update the `videoDataList` by preserving the time points when the subject enters and leaves the lab space where the loading/ unloading task occurs. This list will be used in the following code chunks to only preserve the feature/ profile data corresponding to the continues walking periods.

```{r dataCleaning, dependson="read"}

for (i in seq(length(rawSubjectsData))) {
  videoDataList[[i]] %<>%
    dplyr::select(Leaves, Enters) %>%
    apply(MARGIN=2, FUN=na.omit) %>% # removing NA observations
    as.data.frame %>%
    mutate(Enters = Enters - 600, Leaves = Leaves - 600) %>% # subtracting 600 seconds for warm-up
    filter(Enters > 0 & Leaves > 0) # removing observations/tasks, which contained any of the warm-up period
}

```


## Rational Subgroups
Given the nature of the experimental study, the subjects performed varying tasks that may affect our gait analysis; we have only examined data where the subjects are outside of the main laboratory room since this captures the material handling component of the examined task. An immediate consequence of using only "out-of-the-room" data is that the time differences between the preserved segmented gait cycles is no longer almost constant since relatively large time gaps is introduced to the data that corresponds to the periods of loading/ unloading in the room. For this reason, we have subgrouped the data. Consecutive "out-of-the-room" gait cycles are placed in one subgroup. Statistically speaking, our subgrouping strategy follows the **rational subgrouping** mantra in statistical surveillance. The purpose of rational subgrouping is two-fold:

<!-- Moreover, subgrouping enables us to detect the changes in subgrouped gait cycles instead of individual ones. This results in a more robust monitoring scheme since it can detect more sustained changes. -->

  - Minimize the variability within a subgroup such that the within-group variability captures the natural variability in the gait cycles.
  - Between group variability can be used to capture the effects of "assignable causes of variability", which can include disease progression, fatigue, etc.

For more details on rational subgrouping, we refer the reader to @montgomery2019introduction for more details. The code chunk below provides our approach to creating the rational subgroups for both the "profile" and "gait features" data sets.

```{r subgroups, warning=FALSE}
numRows <- lengths(subjectProfilesList)/5
confidence <- videoDataList %>% lapply(function(x) {mean(x[,2] - x[,1])}) %>%
  do.call(what=c) %>% `*` (0.05) # 5% confidence for each subgroup

subgroupList <- list()
for (id in seq(length(rawSubjectsData))) {

  aM <- subjectProfilesList[[id]] %>% do.call(what=c) %>%
    .[(numRows[[id]]+1):(2*numRows[id])] %>%
    lapply(as.vector)

  expT <- subjectProfilesList[[id]] %>% do.call(what=c) %>%
    .[(4*numRows[[id]]+1):(5*numRows[id])] %>%
    lapply(function(x) {x[1,1]}) %>% do.call(what=c)

  ###########################################################
  # Match times of walking cycles, extracted from videos, against the experimental times
  leaveIdx <- match.closest(videoDataList[[id]]$Leaves + confidence[id], expT)
  enterIdx <- match.closest(videoDataList[[id]]$Enters - confidence[id], expT)

  rmIdx <- c(3000, which(enterIdx - leaveIdx < 5))
  leaveIdx <- leaveIdx[-rmIdx]
  enterIdx <- enterIdx[-rmIdx]

  ########################################################## Create subgroups
  aMList <- list()
  LHDList <- list()
  for (i in 1:length(leaveIdx)) {
    aMList[[i]] <- aM[leaveIdx[i]:enterIdx[i]]
    LHDList[[i]] <- subjectFeaturesList[[id]][leaveIdx[i]:enterIdx[i],]
  }

  ######################################### Remove subgroups with length < 20
  LHDList %<>% .[-{which(lengths(aMList) < 20) %>% c(3000)}]
  aMList %<>% .[-{which(lengths(aMList) < 20) %>% c(3000)}]

  ######################################### Remove observations with profiles< 2nd percentile profile length; This is done to cut all profiles at this length (equal number of features)
  cutLen <- aMList %>% do.call(what=c) %>% lengths() %>% quantile(probs=0.02)
  for (i in 1:length(aMList)) {
    rmIdx <- which(lengths(aMList[[i]]) < cutLen) %>% c(3000)
    aMList[[i]] <- aMList[[i]][-rmIdx]
    LHDList[[i]] <- LHDList[[i]][-rmIdx,]
  }

  ######################################## Create a matrix with cutLen number of columns in each profile subgroup
  aMList %<>% lapply(function(x) {do.call(rbind, x)}) %>%
    lapply(function(x) {x[,1:cutLen]})

  ############################################ Lengths of subgroups 0&1: 10&20
  subgroupList[[id]] <- list(Acc = list(X0=aMList[1:10], X1=aMList[11:30], X2=aMList[31:length(aMList)]),
                             LHD = list(X0=LHDList[1:10], X1=LHDList[11:30], X2=LHDList[31:length(LHDList)]))
}
names(subgroupList) <- c(paste0("subject", c(1:12, 14:15)))

```


# $T^2$ Charting

Using $T^2$ statistic, we compare the mean of subgroups against a reference multivariate distribution by accounting for the correlation between different variables. We calculated the $T^2$ for subgroups using the method introduced in @mason2001applying. Also, to construct a $T^2$ chart, one needs to find the control limits. The control limits are conventionally calculated using the distribution parameters. However, as we investigated the gait data in our analysis, it did not follow any parametric distribution. Therefore, we implemented a non-parametric approach to bootstrap the control limits in both retrospective and prospective phases using the method in @phaladiganon2011bootstrap.


## $T^2$ for Gait Acceleration Profiles

Our $T^2$ charting approach is comprised of 3 phases as described below:

1. **phase 0:** The purpose of phase 0 is to construct a clean *in-control* data set that captures the distribution of the reference data. In this phase we removed the outliers in one pass and updated the subgroups. Then we calculated the within-batch covariance and subgroup means to be used in the following phases.

2. **phase 1:** Although this phase is part of the prospective phase, the control limit for the monitoring phase is calculated using the data in this phase. This is due to the fact that the distribution of the data in retrospective phase differs from that of prospective phase and therefore cannot be directly used for bootstrapping the control limits. To tht end, a set of data from the prospective phase were isolated to be used for bootstrapping the control limit. Note that this set of data were not used for the calculation of the covariance in phase 0.

3. **phase 2:** The $T^2$ was calculated for each prospective subgroup using the mean and covariance parameters, calculated from the phase 0 data. The $T^2$ chart was constructed and the $T^2$ for each subgroup was compared against the control limit. The chart generates an *out-of-control* signal when the $T^2$ is larger than the control limit.

The three phases are implemented in the following code chunk. We present the control charts for the retrospective phase (phase0) and the prospective phase (phase 1 & 2) in the following sections.

```{r T2Profiles}
T2Profiles <- list()
for (id in seq(length(subgroupList))) {
  X0 <- subgroupList[[id]]$Acc$X0

  ############################# Phase0
  ############################# Remove single outliers in number of passes
  pass <- 1
  for (i in 1:pass) {
    UCL0 <- BootstrapUCL0(x0=X0, alpha=0.05, iterations=200) # Find in Functions
    ######################### Find T2 for original X0 and remove outliers from subgroups
    X0Pooled <- do.call(rbind, X0)
    T2 <- X0Pooled %>% apply(FUN=mean, MARGIN=2) %>%
      rep.row(n=nrow(X0Pooled)) %>% -(X0Pooled) %>%
      apply(MARGIN=1, function(x) {x %*% ginv(cov(X0Pooled)) %*% x})

    tmp <- mapply(function(x,y) {rep(y, x)}, x=sapply(X0, nrow), y=1:10) %>% do.call(what=c)
    OCIdx <- T2 %>% split(f=tmp) %>% lapply(function(x) {c(which(x>UCL0), 3000)})
    X0 %<>% mapply(FUN=function(x, y) {x[-y,]}, y=OCIdx) # Updated X0 after phase0
  }

  X0Pooled <- do.call(rbind, X0)
  X0PooledMean <- colMeans(X0Pooled)

  ############################# Within batch covariance; computed from phase0 after outlier removal
  S_w <- X0 %>% lapply(colMeans) %>%
    mapply(FUN=rep.row, sapply(X0, nrow)) %>% # Mean of each subgroup
    mapply(FUN=function(x,y) {x-y}, y=X0, SIMPLIFY=F) %>%
    lapply(function(x) {t(x) %*% x}) %>%
    Reduce(f=`+`) %>%
    `/` (sum(lengths(X0)/ncol(X0[[1]])) - length(X0))

  ############################# Phase1&2
  ############################# Read phase 1&2 subgroups
  X1 <- subgroupList[[id]]$Acc$X1
  X2 <- subgroupList[[id]]$Acc$X2

  ######################### Compute T2 in phase1
  X1Mean <- sapply(X1, colMeans) %>% t()
  T2 <- rep.row(X0PooledMean, length(X1)) %>% -(X1Mean) %>%
    apply(MARGIN=1, function(x) x %*% ginv(S_w) %*% x)

  ######################### Bootstrap UCL using phase 1 data
  sampled <- sample(T2, 1000*length(T2), replace=T) %>% matrix(ncol=length(T2), byrow=T)
  UCL <- apply(sampled, 1, quantile, probs=0.99) %>% mean()

  ######################### T2 for phase2 data
  X12Mean <- sapply(c(X1, X2), colMeans) %>% t()
  T2 <- rep.row(X0PooledMean, length(c(X1, X2))) %>% -(X12Mean) %>%
    apply(MARGIN=1, function(x) x %*% ginv(S_w) %*% x)
  T2Profiles[[id]] <- list(T2=T2, UCL=UCL, UCL0=UCL0)
}
names(T2Profiles) <- paste0("subject", c(1:12, 14:15))
```


### Plots of Phase 0 $T^2$ on Gait Acceleration Profiles along with Reported RPE {.tabset .tabset-fade}

Initialize.


```{r T2Profilesphase0, fig.align='center', results='asis', out.width="100%"}

for (id in seq(length(subgroupList))) {

  X0Pooled <- subgroupList[[id]]$Acc$X0 %>% do.call(what=rbind)
  T2_0 <- rep.row(colMeans(X0Pooled), n=nrow(X0Pooled)) %>% -(X0Pooled) %>%
    apply(MARGIN=1, function(x) {x %*% ginv(cov(X0Pooled)) %*% x})

  UCL0 <- T2Profiles[[id]]$UCL0

  ###############################################################
  ############################################################### T2_0 plot
  T2_0.df <- data.frame(gait=1:length(T2_0), T2_0=T2_0, col=ifelse(T2_0>UCL0, "OC", "IC"))

  T2_0_plot <- ggplot(data=T2_0.df) +
    geom_point(aes(x=gait, y=T2_0, color=col), size=2) +
    scale_colour_manual(values = c("black", "red")) +

    geom_hline(yintercept=UCL0) +
    annotate("text", 2, UCL0, vjust=-1, label="UCL", size=5) +

    ggtitle(paste0("Subject ", id)) +
    xlab("Gait Index") + ylab(expression(T^2 ~ " in Phase 0")) +

    theme_bw() +
    theme(plot.title=element_text(size = 20, hjust=0.5, margin = margin(t=10, b=-30)),
          axis.text=element_text(size=15),
          axis.title=element_text(size=20, color="black"),
          legend.title=element_blank(), # This removes the legend title
          legend.position=c(.1, .85),
          legend.text=element_text(size=15),
          legend.key = element_rect(fill="grey"))

  cat('####',paste0("Subject", id), "{-}",'\n')
  print(T2_0_plot)
  cat('\n \n')
}

```



### Plots of $T^2$ on Gait Acceleration Profiles along with Reported RPE {.tabset .tabset-fade}

Initialize.

```{r T2ProfilesRPEPlot, fig.align='center', results='asis', out.width="100%"}

RPE <- read.csv("https://raw.githubusercontent.com/fmegahed/gait_hotelling/master/Data/csvFiles/RPE/RPE.csv") %>% .[,-14]

SubgroupTimeList <- subgroupList %>% lapply(function(x) {x$LHD[c(2,3)]}) %>%
  lapply(function(x) {do.call(c, x)}) %>%
  lapply(function(x) {sapply(x, function(a) {median(a[,4])})})

for (id in seq(length(subgroupList))) {
  # for (id in 1:2) {
  T2 <- T2Profiles[[id]]$T2
  T2Time <- SubgroupTimeList[[id]]
  UCL <- T2Profiles[[id]]$UCL
  color <- ifelse(T2>UCL, "OC", "IC")
  T2.df <- data.frame(T2Time=T2Time, T2=T2, col=color,
                      shape=c(rep("Phase 1", 20), rep("Phase 2", length(T2) - 20)))
  
  RPE.df <- data.frame(RPETime=RPE[,1], RPE=RPE[,id+1])
  
  ############# Needed to plot the secondary axis
  ylim.prim <- range(pretty(T2))
  ylim.sec <- c(6,20)
  
  b <- diff(ylim.prim)/diff(ylim.sec)
  a <- b*(ylim.prim[1] - ylim.sec[1])
  
  ############# 
  
  T2Plot <- ggplot(data=T2.df) +
    geom_point(aes(x=T2Time, y=T2, shape=shape, col=col), size=2.5) +
    geom_line(aes(x=T2Time, y=T2)) +
    xlab("Experiment Time (s)") + ylab(expression(paste("T"^2, " Profiles"))) +
    
    geom_line(data=RPE.df, mapping=aes(x=RPETime, y=a+RPE*b), color="blue", size=1) +
    # annotate("text", x=RPE.df$RPETime[5], y=(RPE.df$RPE[5]*b*1.1)+a, label="RPE", 
    #        size=5, color="blue") +
    scale_y_continuous(limits=ylim.prim,
                       sec.axis=sec_axis(trans=~ (. - a)/b,
                                       name="RPE Borg Scale", breaks=seq(6,20,2))) +

    geom_hline(yintercept=UCL) +
    annotate("text", 2, UCL, vjust=-1, label="UCL", size=5) +

    scale_shape_manual(values = c(6, 16)) +
    scale_color_manual(values = c("black", "red")) +
    ggtitle(paste0("Subject ", id)) +

    theme_bw() +
    theme(plot.title=element_text(size = 20, hjust=0.5, margin = margin(t=10, b=-30)),
          axis.text=element_text(size=15),
          axis.title=element_text(size=20, color="black"),
          legend.title=element_blank(), # This removes the legend title
          legend.position=c(.1, .68),
          legend.text=element_text(size=10),
          legend.key = element_rect(fill="grey"),
          axis.title.y.right=element_text(color="blue"),
          axis.text.y.right=element_text(color="blue"))

  cat('####',paste0("Subject", id), "{-}",'\n')
  print(T2Plot)
  cat('\n \n')
}
```


## $T^2$ for Gait Features

Initialize.


```{r T2Features}
T2Features <- list()
SWList <- list()
X0Updated <- list()
for (id in seq(length(subgroupList))) {
  X0 <- subgroupList[[id]]$LHD$X0 %>% lapply(function(x) {x=x[,1:3]}) # Last column contains experiment time

  ############################# Phase0
  ############################# Remove single outliers in number of passes
  pass <- 1
  for (i in 1:pass) {
    UCL0 <- BootstrapUCL0(x0=X0, alpha=0.05, iterations=200) # Find in Functions
    ######################### Find T2 for original X0 and remove outliers from subgroups
    X0Pooled <- do.call(rbind, X0)
    T2 <- rep.row(colMeans(X0Pooled), n=nrow(X0Pooled)) %>% -(X0Pooled) %>%
      apply(MARGIN=1, function(x) {x %*% ginv(cov(X0Pooled)) %*% x})

    tmp <- mapply(function(x,y) {rep(y, x)}, x=sapply(X0, nrow), y=1:10) %>% do.call(what=c)
    OCIdx <- T2 %>% split(f=tmp) %>% lapply(function(x) {c(which(x>UCL0), 3000)})
    X0 %<>% mapply(FUN=function(x, y) {x[-y,]}, y=OCIdx) # Updated X0 after phase0
  }
  X0Pooled <- do.call(rbind, X0)
  X0PooledMean <- colMeans(X0Pooled)

  ############################# Within batch covariance; computed from phase0 after outlier removal
  S_w <- X0 %>% lapply(colMeans) %>%
    mapply(FUN=rep.row, sapply(X0, nrow)) %>% # Mean of each subgroup
    mapply(FUN=function(x,y) {x-y}, y=X0, SIMPLIFY=F) %>%
    lapply(function(x) {t(x) %*% x}) %>%
    Reduce(f=`+`) %>%
    `/` (sum(lengths(X0)/ncol(X0[[1]])) - length(X0))

  ############################# Phase1&2
  ############################# Read phase 1&2 subgroups
  X1 <- subgroupList[[id]]$LHD$X1 %>% lapply(function(x) {x=x[,1:3]})
  X2 <- subgroupList[[id]]$LHD$X2 %>% lapply(function(x) {x=x[,1:3]})

  ######################### Compute T2 in phase1
  X1Mean <- sapply(X1, colMeans) %>% t()
  T2 <- rep.row(X0PooledMean, length(X1)) %>% -(X1Mean) %>%
    apply(MARGIN=1, function(x) x %*% ginv(S_w) %*% x)

  ######################### Bootstrap UCL using phase 1 data
  sampled <- sample(T2, 1000*length(T2), replace=T) %>% matrix(ncol=length(T2), byrow=T)
  UCL <- apply(sampled, 1, quantile, probs=0.99) %>% mean()

  ######################### T2 for phase2 data
  X12Mean <- sapply(c(X1, X2), colMeans) %>% t()
  T2 <- rep.row(X0PooledMean, length(c(X1, X2))) %>% -(X12Mean) %>%
    apply(MARGIN=1, function(x) x %*% ginv(S_w) %*% x)

  T2Features[[id]] <- list(T2=T2, UCL=UCL, UCL0=UCL0)
  X0Updated[[id]] <- list(X0=X0)
  SWList[[id]] <- list(S_w=S_w)
}
names(T2Features) <- paste0("subject", c(1:12, 14:15))
names(X0Updated) <- paste0("subject", c(1:12, 14:15))
names(SWList) <- paste0("subject", c(1:12, 14:15))
```


### Plots of Phase 0 $T^2$ on Gait Features along with Reported RPE {.tabset .tabset-fade}

Initialize.

```{r T2Featuresphase0, fig.align='center', results='asis', out.width="100%"}
for (id in seq(length(subgroupList))) {
  
  X0Pooled <- subgroupList[[id]]$LHD$X0 %>% do.call(what=rbind) %>% .[,1:3]
  T2_0 <- rep.row(colMeans(X0Pooled), n=nrow(X0Pooled)) %>% -(X0Pooled) %>%
    apply(MARGIN=1, function(x) {x %*% ginv(cov(X0Pooled)) %*% x})
  
  UCL0 <- T2Features[[id]]$UCL0
  
  ###############################################################
  ############################################################### T2_0 plot
  T2_0.df <- data.frame(gait=1:length(T2_0), T2_0=T2_0, col=ifelse(T2_0>UCL0, "OC", "IC"))
  
  T2_0_plot <- ggplot(data=T2_0.df) +
    geom_point(aes(x=gait, y=T2_0, color=col), size=2) +
    scale_colour_manual(values = c("black", "red")) +
    
    geom_hline(yintercept=UCL0) +
    annotate("text", 2, UCL0, vjust=-1, label="UCL", size=5) +
    
    ggtitle(paste0("Subject ", id)) +
    xlab("Gait Index") + ylab(expression(T^2 ~ " in Phase 0")) +
    
    theme_bw() +
    theme(plot.title=element_text(size = 20, hjust=0.5, margin = margin(t=10, b=-30)),
          axis.text=element_text(size=15),
          axis.title=element_text(size=20, color="black"),
          legend.title=element_blank(), # This removes the legend title
          legend.position=c(.1, .85),
          legend.text=element_text(size=15),
          legend.key = element_rect(fill="grey"))
  
  
  cat('####',paste0("Subject", id), "{-}",'\n')
  print(T2_0_plot)
  cat('\n \n')
  
}
```



### Plots of $T^2$ on Gait Features along with Reported RPE {.tabset .tabset-fade}

Initialize.

```{r T2FeaturesRPEPlot, fig.align='center', results='asis', out.width="100%"}

RPE <- read.csv("https://raw.githubusercontent.com/fmegahed/gait_hotelling/master/Data/csvFiles/RPE/RPE.csv") %>% .[,-14]

SubgroupTimeList <- subgroupList %>% lapply(function(x) {x$LHD[c(2,3)]}) %>%
  lapply(function(x) {do.call(c, x)}) %>%
  lapply(function(x) {sapply(x, function(a) {median(a[,4])})})

for (id in seq(length(subgroupList))) {
  # for (id in 1:2) {
  T2 <- T2Features[[id]]$T2
  T2Time <- SubgroupTimeList[[id]]
  UCL <- T2Features[[id]]$UCL
  color <- ifelse(T2>UCL, "OC", "IC")
  T2.df <- data.frame(T2Time=T2Time, T2=T2, col=color,
                      shape=c(rep("Phase 1", 20), rep("Phase 2", length(T2) - 20)))
  
  RPE.df <- data.frame(RPETime=RPE[,1], RPE=RPE[,id+1])
  
  ############# Needed to plot the secondary axis
  ylim.prim <- range(pretty(T2))
  ylim.sec <- c(6,20)
  
  b <- diff(ylim.prim)/diff(ylim.sec)
  a <- b*(ylim.prim[1] - ylim.sec[1])
  
  ############# 
  
  T2Plot <- ggplot(data=T2.df) +
    geom_point(aes(x=T2Time, y=T2, shape=shape, col=col), size=2.5) +
    geom_line(aes(x=T2Time, y=T2)) +
    xlab("Experiment Time (s)") + ylab(expression(paste("T"^2, " Features"))) +
    
    geom_line(data=RPE.df, mapping=aes(x=RPETime, y=a+RPE*b), color="blue", size=1) +
    # annotate("text", x=RPE.df$RPETime[5], y=(RPE.df$RPE[5]*b*1.1)+a, label="RPE", 
    #        size=5, color="blue") +
    scale_y_continuous(limits=ylim.prim,
                       sec.axis=sec_axis(trans=~ (. - a)/b,
                                       name="RPE Borg Scale", breaks=seq(6,20,2))) +

    geom_hline(yintercept=UCL) +
    annotate("text", 2, UCL, vjust=-1, label="UCL", size=5) +

    scale_shape_manual(values = c(6, 16)) +
    scale_color_manual(values = c("black", "red")) +
    ggtitle(paste0("Subject ", id)) +

    theme_bw() +
    theme(plot.title=element_text(size = 20, hjust=0.5, margin = margin(t=10, b=-30)),
          axis.text=element_text(size=15),
          axis.title=element_text(size=20, color="black"),
          legend.title=element_blank(), # This removes the legend title
          legend.position=c(.1, .68),
          legend.text=element_text(size=10),
          legend.key = element_rect(fill="grey"),
          axis.title.y.right=element_text(color="blue"),
          axis.text.y.right=element_text(color="blue"))

  cat('####',paste0("Subject", id), "{-}",'\n')
  print(T2Plot)
  cat('\n \n')
}
```

# Diagnosis of Gait Features

## Visual Inspection of Profile $T^2$ {.tabset .tabset-fade}

Initialize.

```{r T2ProfilesVisDiag, fig.align='center', results='asis', out.width="100%"}


for (id in seq(length(subgroupList))) {
  
  ##################### Create the acc matrix with ncol=largest length with NA's 
  a <- subjectProfilesList[id][[1]] %>% .[(numRows[id]+1):(2*numRows[id])] %>%
    lapply(unlist) %>% lapply(as.vector)
  
  a %<>% lapply(function(x) {c(x, rep(NA, (max(lengths(a)) - length(x))))}) %>%
    do.call(what=rbind)
  
  ##################### Find cleaned phase 0 indices in acc matrix
  X0Pooled <- subgroupList[[id]]$Acc$X0 %>% do.call(what=rbind)
  T2_0 <- rep.row(colMeans(X0Pooled), n=nrow(X0Pooled)) %>% -(X0Pooled) %>%
    apply(MARGIN=1, function(x) {x %*% ginv(cov(X0Pooled)) %*% x})
  
  UCL0 <- T2Profiles[[id]]$UCL0
  
  phase0Idx <- setdiff(1:nrow(X0Pooled), which(T2_0>UCL0))
  
  ##################### Find indices of observations within subgroup with largest T2 in acc matrix
  expT <- subjectProfilesList[id][[1]] %>% .[(4*numRows[id]+1):(5*numRows[id])] %>%
    sapply(function(x) {x[[1]][1,1]})
  
  subgIdx <- T2Profiles[[id]]$T2[21:length(T2Profiles[[id]]$T2)] %>%
    which.max() # Subgroup with highest T2 in X2 (phase 2)
  LargestT2Times <- subgroupList[[id]]$LHD$X2[[subgIdx]][,4]
  OCIdx <- match.closest(LargestT2Times, expT)
  
  ##################### Plot
  
  cat('###',paste0("Subject", id), "{-}",'\n')
  
  plot(fdata(a[phase0Idx,]), col="grey", ylim=c(-5, 40), xlim=c(0, 70),
     xlab="Gait Time Index", ylab="Acceleration Magnitude",
     cex.axis=1.5, cex.lab=1.5, cex.main=1.5,
     main="Profile Visual Inspection")

  # rect(par("usr")[1],par("usr")[3],par("usr")[2],par("usr")[4],col=gray(0.7, 0.3))
  grid(col="gray")
  lines(fdata(a[OCIdx,]), col="red")
  legend("topleft", legend=c("Phase 0 Profiles Excluding Outliers",
                             expression("Profiles Within the Subgroup with Largest T"^2)),
         lty=1, col=c("grey", "red"), lwd=4, cex=1)
  
  cat('\n \n')
}
```

## Decomposition of Features $T^2$ {.tabset .tabset-fade}

Initialize.


```{r T2Decompose, results='asis'}

qList <- list()
for (id in seq(length(subgroupList))) {
# for (id in 1:2) {
  ############### Read inputs
  X0 <- X0Updated[[id]]$X0
  X1 <- subgroupList[[id]]$LHD$X1 %>% lapply(function(x) {x=x[,1:3]})
  X2 <- subgroupList[[id]]$LHD$X2 %>% lapply(function(x) {x=x[,1:3]})
  X0PooledMean <- do.call(rbind, X0) %>% colMeans()
  X1Mean <- sapply(X1, colMeans) %>% t()
  X12Mean <- sapply(c(X1, X2), colMeans) %>% t()
  S_w <- SWList[[id]]$S_w

  T2 <- T2Features[[id]]$T2
  UCL <- T2Features[[id]]$UCL
  t3 <- which(T2 > UCL)
  p <- 3 # number of features
  alpha <- 0.01

  ############## T2 Decomposition and UCL's
  q <- init.decomp(p=p) # initialize the decomposition matrix
  for(i in 1:nrow(q)) {# Control limits for each component of the decomposition
    b <- subset(q[i,3:ncol(q)], q[i,3:ncol(q)] > 0)

    T2 <- rep.row(X0PooledMean[b], length(X1)) %>% -(X1Mean[,b]) %>%
      apply(MARGIN=1, function(x) x %*% solve(S_w[b,b]) %*% x)

    sampled <- matrix(sample(T2, 1000 * length(T2), replace=T), ncol=20, byrow=T)
    q[i,2] <- apply(sampled, 1, quantile, probs=1-alpha) %>% mean() %>% round(digits=3)
  }

  ############## Decompose T2 for every out of control data point
  cat('###',paste0("Subject", id), "{.tabset .tabset-fade}", '\n')
  
  tmpList <- list()
  for(ii in 1:length(t3)){
    for(i in 1:nrow(q)) {
      b <- subset(q[i,3:ncol(q)], q[i,3:ncol(q)] > 0)
      q[i,1] <- t(X0PooledMean[b]-X12Mean[t3[ii],][b]) %*%
        solve(S_w[b,b]) %*%
        (X0PooledMean[b]-X12Mean[t3[ii],][b]) %>% round(digits=3)
    }
    ############ Store decomposed elements
    tmp <- data.frame(q)
    tmp <- replace(tmp, tmp==1, "Stride Length")
    tmp <- replace(tmp, tmp==2, "Stride Height")
    tmp <- replace(tmp, tmp==3, "Stride Duration")
    tmp <- replace(tmp, tmp==0, "-")
    tmpList[[ii]] <- tmp
    colnames(tmpList[[ii]]) <- c("Decomposition", "Control Limit",
                                 "Selected Feature 1", "Selected Feature 2",
                                 "Selected Feature 3")
    
    cat('####',paste0("subgroup ", t3[[ii]]), "{-}",'\n')
    print(knitr::kable(tmpList[[ii]]))
    cat('\n \n')
    
  }
  
  cat('\n \n')
  
  names(tmpList) <- paste0("subgroup ", t3)
  qList[[id]] <- tmpList
}
names(qList) <- paste0("subject", c(1:12, 14:15))

# write.csv(qList[[1]][[1]], file="C:/Users/Saeb/Desktop/decomposition.csv", quote=F)
# write.table(qList[[1]][[1]], file="C:/Users/Saeb/Desktop/decomposition.txt", quote=F)
```


## 2D Ellipse plots {.tabset .tabset-fade}

Initialize.


```{r Ellipsechart, fig.align='center', results='asis', out.width="100%"}
################ Select subject number and features

for (id in seq(length(subgroupList))) {
  ################ Read inputs
  X0 <- X0Updated[[id]]$X0
  X1 <- subgroupList[[id]]$LHD$X1 %>% lapply(function(x) {x=x[,1:3]})
  X2 <- subgroupList[[id]]$LHD$X2 %>% lapply(function(x) {x=x[,1:3]})
  X0Pooled <- do.call(rbind, X0)
  X0PooledMean <- do.call(rbind, X0) %>% colMeans()
  X1Pooled <- do.call(rbind, X1)
  S_w <- SWList[[id]]$S_w
  q <- qList[[id]][[1]] # Read one of the q's as its 2nd column (q[,2]) are the control limits and thus the same across all subgroups in one subject
  T2 <- T2Features[[id]]$T2
  UCL <- T2Features[[id]]$UCL
  
  cat('###',paste0("Subject", id), "{.tabset .tabset-fade}", '\n')
  
  featurePair <- combn(x=length(X0PooledMean), m=2)
  for (i in seq(ncol(featurePair))) {
    f <- featurePair[,i] # 2 selected features; The features to investigate should be selected by 1st interpreting the T2 decomposition, however here we provide ellipse charts for all combinations
    ################ Univariate CL's and 2D plot limits
    XLim <- c(X0PooledMean[f[1]]-sqrt(S_w[f[1],f[1]]*q[f[1],2]),
              X0PooledMean[f[1]]+sqrt(S_w[f[1],f[1]]*q[f[1],2]))
    
    YLim <- c(X0PooledMean[f[2]]-sqrt(S_w[f[2],f[2]]*q[f[2],2]),
              X0PooledMean[f[2]]+sqrt(S_w[f[2],f[2]]*q[f[2],2]))
    
    xlim <- c(min(XLim) - 1.2*diff(XLim), max(XLim) + 1.2*diff(XLim))
    ylim <- c(min(YLim) - 1.2*diff(YLim), max(YLim) + 1.2*diff(YLim))
    
    ################ Create grid for the ellipse contour plot
    ngrid <- 100
    grid <- cbind(seq(xlim[1], xlim[2], length=ngrid),
                  seq(ylim[1], ylim[2], length=ngrid))
    
    ################  T2 on the grid points
    T2_grid <- apply(expand.grid(grid[,1], grid[,2]), 1,
                     function(x) {(x- X0PooledMean[f]) %*% solve(S_w[f,f]) %*% (x- X0PooledMean[f])})
    T2_grid <- matrix(T2_grid, ngrid, ngrid)
    
    cat('####',paste("Features", f[1], "&", f[2]), "{-}",'\n')
    ######################################################## Plot
    ############## Plot the contour
    # old.par <- par()
    
    # par(mar=c(5, 5, 4, 2) + 0.1)
    
    plot(1 ,1, type = "n", xlim = xlim, ylim = ylim, main="Ellipse Chart",
         xlab=paste0(colnames(X0[[1]])[f[1]], " (m)"),
         ylab=paste0(ylab=colnames(X0[[1]])[f[2]], " (s)"),
         cex.axis=1.5, cex.lab=1.5, cex.main=1.5)
    
    grid()
    
    legend("topleft", legend=c("phase 0 subgroup means", "phase 1 subgroup means",
                               "X control limits", "Y control limits"),
           col=c("darkgreen", "blue", "black", "brown"), pch=c(3,3,NA,NA), lty=c(NA,NA,2,2),
           cex=1, pt.cex=1.2, lwd=2, bg=gray(0.9), y.intersp=0.8)
    ################
    tmp <- c("Stride Length", "Stride Height", "Stride Duration")[f]
    idx <- apply(q[,3:5], MARGIN=1, function(r) all(c(tmp, "-") %in% r)) %>% which()
    contour(grid[,1], grid[,2], T2_grid, levels=q[idx,2], drawlabels=T, add=T, labcex=1)
    
    ################ Plot phase 0 data
    center0 <- X0 %>% lapply(function(x) {x[,f]}) %>% sapply(colMeans) %>% t()
    points(center0[,1], center0[,2], pch=3, cex=1.5, col="darkgreen", lwd=2)
    # points(X0Pooled[,f], pch=1, cex=0.5, col="blue")
    
    ################ Plot phase 1 data
    center1 <- X1 %>% lapply(function(x) {x[,f]}) %>% sapply(colMeans) %>% t()
    points(center1[,1], center1[,2], pch=3, cex=1.5, col="blue", lwd=2)
    # points(X1Pooled[,f], pch=1, cex=0.5, col="blue")
    
    ################ plot phase 1&2 data by subgroup number
    center12 <- c(X1,X2) %>% lapply(colMeans) %>% sapply(function(x) {x[f]}) %>% t()
    labels <- T2 %>% t() %>% as.data.frame() %>% names() %>% gsub(pattern='V', replacement='')
    text(center12, labels=labels, cex = 0.7, col=ifelse(T2>UCL, "red", "black"))
    
    ################ Plot univariate CL's
    abline(v=XLim, lty="dashed", col="black", lwd=2)
    abline(h=YLim, lty="dashed", col="brown", lwd=2)
    
    cat('\n \n')
  }
  
  cat('\n \n')
}

```



---

# References {-}
